Results publications are inadequately linked to trial registrations: An automated pipeline and evaluation of German university medical centers

### 2021-07-26 docs/manuscript.Rmd
Update manuscript with additional analyses and papaja

Clinical trial results risk being left out of evidence synthesis and health policy decision making if publications cannot be found. Incomplete evidence threatens healthcare and the lives and welfare of human volunteers. Legal regulations as well as ethical guidelines call for trial results transparency [@worldmedicalassociation2013; @worldhealthorganization2017; @altman2014gfrhraum; @borysowski2020bjcp]. Linking trial registrations and results publications makes each more findable, reducing the risk of information loss and increasing the clinical research value[@kubben2019] <!--# looked for pub on FAIR/finable clinical trials and not finding much beyond just data-->. Furthermore, such threaded evidence empowers readers to cross-check sources, e.g., for potential outcome switching [@altman2014t], and can support systematic reviewers who rely on registrations to supply additional trial information for evidence synthesis [@miron2020sd].

@@

## Trial and publication characteristics

@@

## Registration-publication links

@@

### Logistic regression

@@

Table \@ref(tab:tbl-trials-regression) shows the crude univariate and adjusted multivariate odds ratios (ORs) for each type of publication-registration link across all explanatory variables. Completion year did not have a strong relationship with linking practices, although trials completing more recently were significantly more likely report the TRN in the abstract (`r inline_text(cor_abs, variable = "completion_year", pattern = "cOR {estimate} [{ci}]")`; `r inline_text(aor_abs, variable = "completion_year", pattern = "aOR {estimate} [{ci}]")`) and full-text (`r inline_text(cor_ft, variable = "completion_year", pattern = "cOR {estimate} [{ci}]")`; `r inline_text(aor_ft, variable = "completion_year", pattern = "aOR {estimate} [{ci}]")`). Figure \@ref(fig:plot-links-year-reg) shows the rate of registration-publication links over time for ClinicalTrials.gov and DRKS.

@@

With regards to the relationship between link types,<!--#in the adjusted analysis,--> trials with one type of link were generally more likely to have another type of link. <!--#However, trials with a TRN in the full-text were significantly less likely to have the publication linked in the registration (`r inline_text(aor_reg, variable = "has_iv_trn_ft_pdf", pattern = "aOR {estimate} [{ci}]")` ; `r inline_text(aor_ft, variable = "has_reg_pub_link", pattern = "aOR {estimate} [{ci}]")`).--> In particular, TRNs were significantly more likely to appear in the PubMed metadata if included in the abstract (`r inline_text(aor_si, variable = "has_iv_trn_abstract", pattern = "aOR {estimate} [{ci}]")`), but not significantly more likely if included in the full-text (`r inline_text(aor_si, variable = "has_iv_trn_ft_pdf", pattern = "aOR {estimate} [{ci}]")`). Futhermore, DRKS TRNs were significantly less likely than ClinicalTrials.gov TRNs to appear in the PubMed metadata (`r inline_text(aor_si, variable = "registry", level = "DRKS", pattern = "aOR {estimate} [{ci}]")`). Similarly, for trials with a TRN in either abstract or full-text, ClinicalTrials.gov TRNs appeared at a higher rate than DRKS TRNs in the metadata (62% vs. 18%). Table X shows the number and proportion of trials with a TRN in the metadata given TRN in the abstract, full-text, or either, overall and by registry.

@@

### Publications linked in registrations

We found that `r report_pct_num_denom_trials(n_trials_link_reg)` of registrations included a reference with a publication identifier (DOI or PMID) to the associated publication in our sample of trials with published results. Only `r report_pct_num_denom_trials(n_trials_link_reg_drks, n_trials_drks)` of DRKS trials referenced the results publication in the registration, in contrast with `r report_pct_num_denom_trials(n_trials_link_reg_ctgov, n_trials_ctgov)` of ClinicalTrials.gov trials, of which `r report_pct_num_denom_trials(n_trials_ctgov_link_reg_trn_si_derived, n_trials_link_reg_ctgov)` were automatically indexed from PubMed metadata.

<!--# @daniel: you had several comments about automatically linked pubs so I don't think the language I had was clear. I tried several wordings below. Let's discuss and figure out what is clearest. -->

Of ClinicalTrials.gov trials with the TRN in the PubMed metadata (n = `r n_trials_ctgov_link_reg_trn_si`), `r report_n_pct_trials(n_trials_ctgov_link_reg_trn_si_manual + n_trials_ctgov_no_link_reg_trn_si, n_trials_ctgov_link_reg_trn_si)` were not automatically indexed.

Of ClinicalTrials.gov trials with the TRN in the PubMed metadata (n = `r n_trials_ctgov_link_reg_trn_si`), `r report_n_pct_trials(n_trials_ctgov_link_reg_trn_si_derived, n_trials_ctgov_link_reg_trn_si)` have the publication automatically linked in the registry, `r report_n_pct_trials(n_trials_ctgov_link_reg_trn_si_manual, n_trials_ctgov_link_reg_trn_si)` have the publication manually linked in the registry, and only `r report_n_pct_trials(n_trials_ctgov_no_link_reg_trn_si, n_trials_ctgov_link_reg_trn_si)` trial is not linked in the registry.

All `r n_trials_ctgov_link_reg_trn_si_derived` trials with automatically indexed publications in ClinicalTrials.gov include the TRN in the PubMed metadata. An additional `r n_trials_ctgov_link_reg_trn_si_manual` trials include the TRN in the PubMed metadata and a link publication that was not automatically indexed, as it may have been entered by the trialist manually prior to being indexed. `r if (n_trials_ctgov_no_link_reg_trn_si == 1) "Only one"` ClinicalTrials.gov trial (NCT01614301) has a TRN in the PubMed metadata (PMID: 26417987) and no linked publication in the registration. While the TRN is indeed indexed in the secondary identifier field, PubMed mistakenly associated it with an incorrect databank (i.e., GENBANK) which presumably prevented ClinicalTrials.gov from making the link.

When limiting to only manually linked publications in the registration, performance is similar between ClinicalTrials.gov (24%, 186/783) and DRKS (22%, 96/447). <!--# change to code -->

### Registrations linked in publications

We found that `r report_pct_num_denom_trials(n_trials_link_trn_none)` trials did not report the known TRN in any of the metadata, abstract, or full-text of the corresponding publication. Only `r report_pct_num_denom_trials(n_trials_link_trn_all)` reported the TRN across all three locations. TRNs were most commonly reported in the full-text, appearing in `r report_n_pct_trials(n_trials_link_trn_ft)`. TRNs were included in `r report_n_pct_trials(n_trials_link_trn_abs)` of abstracts, as compared with `r report_n_pct_trials(n_trials_link_trn_si)` of PubMed metadata.

### Linking by registry

We found that trials with a ClinicalTrials.gov TRN were significantly more likely than trials with a DRKS TRN to include a publication reference in the registry (`r apa::apa(chisq_link_by_reg_reg, print_n = TRUE)`) and to include the TRN in the PubMed metadata (`r apa::apa(chisq_link_by_reg_si, print_n = TRUE)`) and abstract (`r apa::apa(chisq_link_by_reg_abs, print_n = TRUE)`). We did not find a significant relationship between registry and TRN reporting in full-text (`r apa::apa(chisq_link_by_reg_pdf, print_n = TRUE)`).

### TRN reporting in PubMed metadata

<!--# @daniel asked me to test relationship between trn in metadata and abs/ft. many options. total in abs 714 and ft 1131. chisquare trn reporting abs given diff combos of abs/ft reporting OR just abs or ft? ideally, compare registry, trn in abs/ft (not sure what levels), and trn in si, but 3 variable so not chisquare-->

For the 1,372 trials that provided a TRN in either the abstract or full-text, a TRN was also available in the metadata in 55% (n = 767). ClinicalTrials.gov TRNs (65%, 705/1077) were more likely than DRKS TRNs (21%, 62/295) to be included in PubMed metadata. Furthermore, 97% (563/580) of trials with a ClinicalTrials.gov TRN in the abstract also had the TRN in the PubMed metadata, compared with 23% (31/134) of trials with a DRKS TRN. Similarly, 58% (507/878) of trials with a ClinicalTrials.gov TRN in the full-text also had the TRN in the PubMed metadata, compared with 18% (48/253) of trials with a DRKS TRN.

A chi-square test of independence was performed to examine the relation between registry and TRN in PubMed metadata, for trials with TRN reported in both the abstract and the full-text. The relation between these variables was significant, X2 (1, N = 473) = 201.0, p = 1.07e-45. Trials with a ClinicalTrials.gov TRN (Proportion with TRN in SI = 95.5%) were more likely than trials with a DRKS TRN (Proportion with TRN in SI = 33.7%) to have a TRN in PubMed metadata if reported in both the abstract and the full-text (Difference in reporting = 61.8%, CI = 52.5, 71.4).

## Cross-registrations

We found `r report_pct_num_denom_trials(n_trials_crossreg_reg)` of trials listed additional TRNs in their registrations, indicating likely cross-registrations. Of these, 18 trials (6%) listed more than 1 cross-registration, for a total of `r n_crossreg_reg` cross-registrations. Of these `r n_crossreg_reg` cross-registrations, X were reported in pubs (X in metadata, X in abs, X in ft). Beyond the IntoValue primary TRNs and these registry-indicated cross-registrations, we detected X additional TRNs in X publications. Table \@ref(tab:tbl-crossreg) shows details on cross-registrations and additional TRNs by registry.

```{r tbl-crossreg}
tbl_crossreg %>% 
  gtsummary::as_flex_table() %>% 
  flextable::set_caption(caption = "Cross-registrations and additional TRNs by registry")
```

@@

We detected X TRNs in PubMed metadata, abstracts, and full-text <!--# Some trial publications also reported TRNs --> which did not correspond to either the associated trial's IntoValue TRN or a cross-registration listed on the registry. While the TRNs in the PubMed metadata (n = X) and abstracts (n = X) are likely supplemental cross-registrations, the additional TRNs in the full-text may also have been provided as background or discussion. Such discrimination falls beyond the scope of these regexes.

@@

[TO ADD]

>>>>

Clinical trial results risk being left out of evidence synthesis and health policy decision making if publications cannot be found. Incomplete evidence threatens healthcare and the lives and welfare of human volunteers. Legal regulations as well as ethical guidelines call for trial results transparency [@worldmedicalassociation2013; @worldhealthorganization2017; @altman2014gfrhraum; @borysowski2020bjcp]. Linking trial registrations and results publications makes each more findable, reducing the risk of information loss and increasing the clinical research value[@kubben2019]<!--# looked for pub on FAIR/findable clinical trials and not finding much beyond just data-->. Furthermore, such threaded evidence empowers readers to cross-check sources, e.g., for potential outcome switching [@altman2014t], and can support systematic reviewers who rely on registrations to supply additional trial information for evidence synthesis [@miron2020sd].

@@

@@

@@

Table \@ref(tab:tbl-trials-regression) shows the crude univariate and adjusted multivariate odds ratios (ORs) for each type of publication-registration link across all explanatory variables.

@@

Completion year did not have a strong relationship with linking practices, although trials completing more recently were significantly more likely report the TRN in the abstract (`r inline_text(cor_abs, variable = "completion_year", pattern = "cOR {estimate} [{ci}]")`; `r inline_text(aor_abs, variable = "completion_year", pattern = "aOR {estimate} [{ci}]")`) and full-text (`r inline_text(cor_ft, variable = "completion_year", pattern = "cOR {estimate} [{ci}]")`; `r inline_text(aor_ft, variable = "completion_year", pattern = "aOR {estimate} [{ci}]")`). Figure \@ref(fig:plot-links-year-reg) shows the rate of registration-publication links over time for ClinicalTrials.gov and DRKS.

@@

Across the adjusted multivariate models,<!--#in the adjusted analysis,--> trials with one type of link were generally more likely to have another type of link. <!--#However, trials with a TRN in the full-text were significantly less likely to have the publication linked in the registration (`r inline_text(aor_reg, variable = "has_iv_trn_ft_pdf", pattern = "aOR {estimate} [{ci}]")` ; `r inline_text(aor_ft, variable = "has_reg_pub_link", pattern = "aOR {estimate} [{ci}]")`).--> In particular, TRNs were significantly more likely to appear in the PubMed metadata if included in the abstract (`r inline_text(aor_si, variable = "has_iv_trn_abstract", pattern = "aOR {estimate} [{ci}]")`), but not significantly more likely if included in the full-text (`r inline_text(aor_si, variable = "has_iv_trn_ft_pdf", pattern = "aOR {estimate} [{ci}]")`). Futhermore, DRKS TRNs were significantly less likely than ClinicalTrials.gov TRNs to appear in the PubMed metadata (`r inline_text(aor_si, variable = "registry", level = "DRKS", pattern = "aOR {estimate} [{ci}]")`). Similarly, for trials with a TRN in either abstract or full-text, ClinicalTrials.gov TRNs appeared at a higher rate than DRKS TRNs in the metadata (62% vs. 18%). Table X shows the number and proportion of trials with a TRN in the metadata given TRN in the abstract, full-text, or either, both overall and by registry.

@@

@@

@@

We would like to thank Kerstin Rubarth from the Charité Institut für Biometrie und Klinische Epidemiologie for her consultation on the statistical analysis.
